---
title: "Introduction to Statistics"
author: "Nicola Rennie"
format:
  revealjs:
    resources:
    - webr-serviceworker.js
    - webr-worker.js
webr:
  channel-type: "automatic"
  packages: ['ggplot2', 'dplyr']
filters:
- webr
---

# Descriptive statistics {background-color="#1A936F"}

Descriptive statistics provide a summary that quantitatively describes a sample of data.

```{r}
#| label: setup
#| echo: false
#| eval: true
#| message: false
library(tidyverse)
library(emojifont)
library(showtext)
library(reactable)
library(kableExtra)
font_add_google("Ubuntu", "Ubuntu")
showtext_auto()
set.seed(1234)
population_df = tibble(ID = 1:200, 
                       x = rep(1:20, times = 10),
                       y = rep(1:10, each = 20),
                       Value = rnorm(200, 250, 50))
sample_size = 10
sample_ids = sample(1:200, size = sample_size, replace = FALSE)
sample_df =  filter(population_df, ID %in% sample_ids)
```

## Population

**Population** refers to the entire group of individuals that we want to draw conclusions about. 

```{r}
#| label: pop-people
#| eval: true
#| echo: false
#| fig-align: center
#| fig-height: 4.16
ggplot() +
  geom_text(data = population_df,
            mapping = aes(x = x,
                          y = y,
                          label = fontawesome('fa-user'),
                          colour = Value),
            family='fontawesome-webfont', size = 20) +
  scale_colour_gradient(low = "#baded3", high = "#12664d") +
  labs(title = "Population: 200 people") +
  theme_void() +
  theme(legend.position = "none", 
        legend.title = element_blank(),
        plot.margin = margin(10, 10, 10, 10), 
        plot.title = element_text(face = "bold",
                                  hjust = 0.5,
                                  family = "Ubuntu",
                                  size = 36,
                                  margin = margin(b = 10)))

```

## Sample

**Sample** refers to the (usually smaller) group of people for which we have collected data on.

```{r}
#| label: samp-people
#| eval: true
#| echo: false
#| fig-align: center
#| fig-height: 4.16
ggplot() +
  geom_text(data = population_df,
            mapping = aes(x = x,
                          y = y,
                          label = fontawesome('fa-user')),
            family='fontawesome-webfont', size = 20, colour = "grey") +
  geom_text(data = sample_df,
            mapping = aes(x = x,
                          y = y,
                          label = fontawesome('fa-user'),
                          colour = Value),
            family='fontawesome-webfont', size = 20) +
  scale_colour_gradient(low = "#baded3", high = "#12664d") +
  labs(title = glue::glue("Sample: {sample_size} people")) +
  theme_void() +
  theme(legend.position = "none", 
        legend.title = element_blank(),
        plot.margin = margin(10, 10, 10, 10), 
        plot.title = element_text(face = "bold",
                                  hjust = 0.5,
                                  family = "Ubuntu",
                                  size = 36,
                                  margin = margin(b = 10)))

```

## Generate sample data {.scrollable}

For the examples later, let's create a population of data in R...:

```{webr-r}
# Generate population data
set.seed(1234)
population = rnorm(200, 250, 50)
print("Population generated!")
```

## Generate sample data {.scrollable}

... and draw a sample from it:

```{webr-r}
# Pick a sample
set.seed(1234)
sample_size = 10
sample_data = sample(population,
                     size = sample_size,
                     replace = FALSE)
print("You've created a sample of data!")
```

## Sample mean

The mean, often simply called the *average*, is defined as *the sum of all values divided by the number of values*. It's a measure of central tendency that tells us what's happening near the middle of the data.

::::{style='text-align: center;'}

$\bar{x} = \frac{1}{n} \sum_{i=i}^{n} x_{i}$

::::

## Sample mean

In R, we use the `mean()` function:

```{webr-r}
# Calculate mean
mean(sample_data)
```

## Sample standard deviation

The standard deviation tells us about how spread out the data is. A lower standard deviation indicates that values tend to be close to the mean, and a higher standard deviation indicates that the values are spread out over a wider range.

::::{style='text-align: center;'}

$s = \sqrt{\frac{\Sigma_{i= 1}^{N} (x_i - \bar{x})^2}{n-1}}$

::::

## Sample standard deviation

In R, we use the `sd()` function:

```{webr-r}
# Calculate standard deviation
sd(sample_data)
```
